---
title: "Data Analysis & Modelling"
date: "05/16/2020"
output: html_document
---
```{r}
data <- read.csv("data.csv")

library(tidyverse)
library(broom)
library(gtools)
```

# MODELS

## Model 1: SIMPLE MODEL
```{r}
model_simple <- glm(chronic ~ HRSLEEP + I(HRSLEEP^2) + depfreq_often + depfreq_monthly, data = data, family = binomial)

summary(model_simple)

# significant coefficients: all

# Results for simple model:
# sleep significant at 0.001 level, p-value = 0.000447
# sleep squared significant at 0.001 level, p-value = 2.05e-06
# depfreq_often significant at 0.001 level, p-value < 2e-16
# depfreq_monthly significant at 0.05 level, p-value = 0.011191
```

## Model 2: FULL MODEL
```{r}
# New with transformed sleep, excluding interaction with the original sleep variable
# AIC score: 3297.7

model_full <- glm(chronic ~ HRSLEEP + I(HRSLEEP^2) + depfreq_often + depfreq_monthly + I(I(HRSLEEP^2)*depfreq_often) + I(I(HRSLEEP^2)*depfreq_monthly) + log(AGE) + SEX + BMI + activity_score + income_high + income_middle + race_asian + race_black + race_na, family = binomial, data = data)

summary(model_full)

# significant coefficients: sleep, sleep squared, depfreq_often, all covariates

# Results for full model:
# sleep significant at 0.05 level, p-value = 0.0317
# sleep squared significant at 0.05 level, p-value = 0.0149
# depfreq_often significant at 0.001 level, p-value = 1.40e-05
```

## Model 3: FULL WOMEN
```{r}
# Data 1: only women
data1 <- data %>% 
  filter(SEX == 1)

model_women <- glm(chronic ~ HRSLEEP + I(HRSLEEP^2) + depfreq_often + depfreq_monthly + I(I(HRSLEEP^2)*depfreq_often) + I(I(HRSLEEP^2)*depfreq_monthly) + log(AGE) + BMI + activity_score + income_high + income_middle + race_asian + race_black + race_na, family = binomial, data = data1)

summary(model_women)

# significant coefficients: depfreq_often, all covariates except race_na

# Results for women:
# depfreq_often significant at 0.01 level, p-value = 0.00145
```

## Model 4: FULL MEN
```{r}
# Data 2: only men
data2 <- data %>% 
  filter(SEX == 0)

model_men <- glm(chronic ~ HRSLEEP + I(HRSLEEP^2) + depfreq_often + depfreq_monthly + I(I(HRSLEEP^2)*depfreq_often) + I(I(HRSLEEP^2)*depfreq_monthly) + log(AGE) + BMI + activity_score + income_high + income_middle + race_asian + race_black + race_na, family = binomial, data = data2)

summary(model_men)

# significant coefficients: sleep, sleep squared, depfreq_often, most covariates

# Result for men:
# sleep significant at 0.05 level, p-value = 0.02085
# sleep squared significant at 0.05 level, p-value = 0.01464
# depfreq_often significant at 0.01 level, p-value = 0.00292
```

```{r}
# Original without transformed sleep
# AIC score: 3300.2
model_og <- glm(chronic ~ HRSLEEP + depfreq_often + depfreq_monthly + I(HRSLEEP*depfreq_often) + I(HRSLEEP*depfreq_monthly) + log(AGE) + SEX + BMI + activity_score + income_high + income_middle + race_asian + race_black + race_na, family = binomial, data = data)

summary(model_og)
```

## Checking Regression Assumptions: 

# Linearity

The empirical logit plot seems approximately linear.

 ```{r, warning = FALSE, message = FALSE}
### NOT SURE IF WE WANT TO CHANGE THIS
# Code from 18-simplelogistic-lab activity to generate x and y values for plot to check for linearity
data_bin <- data %>%
  # Break HRSLEEP into 10 bins
  mutate(HRSLEEP_group = cut(HRSLEEP, breaks=10)) %>% 
  group_by(HRSLEEP_group) %>%
  # calculate mean HRSLEEP value and proportion obese in each bin
  mutate(binned.y = mean(chronic), binned.x = mean(HRSLEEP)) %>% 
  # transform binned.y from proportion to log(odds)
  mutate(binned.y_logit = logit(binned.y))

# Create plot of log(odds) verus age
ggplot(data_bin, aes(x = binned.x, y = binned.y_logit)) +
  geom_point() + 
  geom_smooth(method = lm, se = FALSE)
```
## Randomness

Obviously, individuals are not assigned BMIs by a spinner model, but since we have a random sample, the randomness assumption is met.

## Independence

As established above for the linear regression model, the independence assumption is reasonable for this data set.

```{r}
data.cor = cor(data)
```

## Likelihood Ratio Tests

$H_0: \beta_i = 0$ 

$H_a: \beta_i \ne 0$

$\alpha = 0.05$

```{r}
# Load necessary package
require(lmtest)

# Full model
lrtest(model_full)
```

The $-2log(L)$ values are $-2\cdot -2012.5 = 4025$ and $-2\cdot -1635.1 = 3270.2$ for the null model and full logistic model, respectively. This yields a G-statistic of 24308 - (23600) = 754.8. Compared to a chi-distribution with 14 degrees of freedom, the p-value is less than 2.2e-16, which is significant at the alpha = 0.05 level. We therefore reject the null hypothesis and conclude that the multiple logistic regression model based on sleep, depression frequency, and the interaction between sleep and depression frequency is effective at explaining the probability of being diabetic.

```{r}
## WHAT IS THIS??
# code from Logistic Regression visualization practice activity 4/20/20
preds_full <- augment(x=model_full, type.predict = "response") %>%
  mutate(lci=`.fitted`-(1.96*`.se.fit`),
         uci=`.fitted`+(1.96*`.se.fit`)) %>%
  rename(prob=`.fitted`)

probs_full<- preds_full %>%
  select(prob,lci,uci, depfreq_often, depfreq_monthly, HRSLEEP, chronic)

library(InformationValue)
Concordance(probs_full$chronic, probs_full$prob)
```
```{r}
# women
lrtest(model_women)
```

```{r}
# code from Logistic Regression visualization practice activity 4/20/20
preds_women <- augment(x=model_women, type.predict = "response") %>%
  mutate(lci=`.fitted`-(1.96*`.se.fit`),
         uci=`.fitted`+(1.96*`.se.fit`)) %>%
  rename(prob=`.fitted`)

probs_women <- preds_women %>%
  select(prob,lci,uci, depfreq_often, depfreq_monthly, HRSLEEP, chronic)

library(InformationValue)
Concordance(probs_women$chronic, probs_women$prob)
```

```{r}
# men
lrtest(model_men)
```

```{r}
# code from Logistic Regression visualization practice activity 4/20/20
preds_men <- augment(x=model_men, type.predict = "response") %>%
  mutate(lci=`.fitted`-(1.96*`.se.fit`),
         uci=`.fitted`+(1.96*`.se.fit`)) %>%
  rename(prob=`.fitted`)

probs_men <- preds_men %>%
  select(prob,lci,uci, depfreq_often, depfreq_monthly, HRSLEEP, chronic)

library(InformationValue)
Concordance(probs_men$chronic, probs_men$prob)
```

```{r}
# no confounders
lrtest(model_simple)
```

```{r}
# code from Logistic Regression visualization practice activity 4/20/20
preds_simple <- augment(x=model_simple, type.predict = "response") %>%
  mutate(lci=`.fitted`-(1.96*`.se.fit`),
         uci=`.fitted`+(1.96*`.se.fit`)) %>%
  rename(prob=`.fitted`)

probs_simple <- preds_simple %>%
  select(prob,lci,uci, depfreq_often, depfreq_monthly, HRSLEEP, chronic)

library(InformationValue)
Concordance(probs_simple$chronic, probs_simple$prob)
```
