---
title: "Data Appendix"
author: "Clara Li and Rose Porta"
date: "5/2/2020"
output: html_document
---
```{r}
# NOTE: To load data, you must download both the extract's data and the DDI
# and also set the working directory to the folder with these files (or change the path below).

if (!require("ipumsr")) stop("Reading IPUMS data into R requires the ipumsr package. It can be installed using the following command: install.packages('ipumsr')")

ddi <- read_ipums_ddi("nhis_00004.xml")
data <- read_ipums_micro(ddi)

# Load necessary packages
library(base)
library(mosaic)
library(tidyverse)
```

# Initial Data Wrangling

```{r}
# Create activity level variable and filter out non-applicable values
data1 <- data %>% 
  mutate( activity_score = (9 * VIG10FWK + 5 * MOD10FWK + 3 * STRONGFWK)) %>% 
  filter(AGE >= 18, 
         BMI != "0", BMI != "0.00", BMI != "99.80", BMI != "99.99", 
         INCFAM07ON != "99", INCFAM07ON != "96", 
         RACEA != "900", RACEA != "970", RACEA != "980", RACEA != "990", RACEA != "600",
         HEALTH != "0", HEALTH != "7", HEALTH != "8", HEALTH != "9", 
         STRONGFWK != "00", STRONGFWK != "96", STRONGFWK != "97", STRONGFWK != "98", STRONGFWK != "99", 
         MOD10FWK != "00", MOD10FWK != "96", MOD10FWK != "97", MOD10FWK != "98", MOD10FWK != "99", 
         VIG10FWK != "00", VIG10FWK != "96", VIG10FWK != "97", VIG10FWK != "98", VIG10FWK != "99",
         HRSLEEP != "00", HRSLEEP != "97", HRSLEEP != "98", HRSLEEP != "99")

View(data1)

data2 <- data1 %>% 
  filter(DEPFREQ != "0", DEPFREQ != "7", DEPFREQ != "8", DEPFREQ != "9",
         FLDIABETNO != "00", FLDIABETNO != "97", FLDIABETNO != "98", FLDIABETNO != "99",
         FLDIABETTP != "7", FLDIABETTP != "8", FLDIABETTP != "9") %>% 
  mutate(SEX = ifelse(SEX == 1, 0, 1))

View(data2)


```

# Variable Analysis

```{r}
# Look at distributions of variables
favstats(~BMI, data = data)
```

The minimum BMI is 16.16 and the maximum is 69.88. There are no missing values. 

```{r}
favstats(~HRSLEEP, data = data)
```

The minimum value for hours of sleep is 2, and the maximum is 18. These both seem like very extreme values, but Q1 is 6 and Q3 is 8, which indicates that most individuals get between 6 and 8 hours of sleep, which makes sense. There are no missing values, which is ideal.

```{r}
favstats(~AGE, data = data)
```

The minimum age is 32 years and the maximum is 85 years. There are no missing values, which is ideal.

```{r}
favstats(~STRONGFWK, data = data)
```

```{r}
favstats(~MOD10FWK, data = data)
```

```{r}
favstats(~VIG10FWK, data = data)
```

For each of the 3 above variables that we used to create our activity_score variable, the minimum is 1 and the maximum is 95. The minimum makes sense, but it seems very extreme and almost impossible for someone to exercise 95 times per week. There are no missing values.

```{r}
favstats(~activity_score, data = data)
```

We created the activity_score variable using the Oncology Nursing Society formula Weekly leisure activity score = (9 × Strenuous) + (5 × Moderate) + (3 × Light) from the Godin Leisure-Time Exercise Questionnaire. We substituted strength exercise for light exercise due to the availability of the IPUMS data, and we think strength training is analagous to light exercise for the purposes of our analysis. Each of the 3 variables used to create this are measured in number of times per week. The minimum value is 17 and the maximum is 1615. We are not sure exactly how to make sense of these numbers, but we are assuming that a higher score means a higher overall activity level. There are no missing values. 

```{r}
tally(~SEX, data = data)
```

The number of females and males is roughly equal (with slighly more females), which makes sense.

```{r}
tally(~INCFAM07ON, data = data)
```

Each category corresponds to a family income-level category, with 10 being the lowest category and 24 being the highest. We thought it was necessary to collapse the categories into low, middle, and high income in order to avoid over-complicating our analysis. We noticed that there are very few people in the lowest category, but the greatest number of people in the second-lowest. Based on this distribution, we decided to combine the two lowest categories into the low income category, the 3 middle categories into the middle income category, and the two highest categories into the high income category. 

```{r}
tally(~RACEA, data = data)
```

Each category code corresponds to a race. We noticed that the vast majority of individuals are white (category 100). 

```{r}
tally(~HEALTH, data = data2)
```

1 corresponds to very good health and 5 corresponds to very poor health. We noticed that most people are in good or excellent health, and fewer people are in poor health. In order to simplify our analysis, we decided to collapse health into a binary variable: good health versus poor health. We combined the 1, 2, and 3 categories into good health (because the middle (3) category was coded in IPUMS as "good"), and the other two into poor health.

```{r}
tally(~DEPFREQ, data = data1)
```

1 corresponds to the highest worry frequency (daily), and 5 corresponds to the lowest. We noticed that there are fewer people who worry a lot, and more who do not report worrying a lot. We think we will need to collapse some of the categories, but since it is not obvious from the tally which ones to combine, we will look at a faceted scatter plot using worry frequency as the explanatory and weight as the response in order to see if particular categories show particularly strong associations with weight. 

```{r}
favstats(~FLDIABETNO, data = data1)
```

# Additional Data Wrangling

```{r}
# Collapse categorical variables
data <- data %>% 
  # income_low is reference group
  mutate(income_high = ifelse(INCFAM07ON == "23" |  INCFAM07ON == "24", 1, 0),
         income_middle = ifelse(INCFAM07ON == "12" | INCFAM07ON == "21" | INCFAM07ON == "22", 1, 0)) %>% 
  # health_binary = 1 indicates good health
  mutate(health_binary = ifelse(HEALTH == "1" | HEALTH == "2" | HEALTH == "3", 1, 0)) %>% 
  select(WEIGHT, HRSLEEP, AGE, BMI, activity_score, SEX, RACEA, 
         income_high, income_middle, health_binary, WORFREQ)
```

# Structure and Names

```{r}
str(data, give.attr = FALSE)
```

After some initial data wrangling, our data set has 11 variables and 19925 observational units. The variables are:

1. WEIGHT is a number indicating the person's weight in pounds
2. HRSLEEP is an integer indicating self-reported number of hours of sleep per night an individual gets.
3. AGE is a number indicating the person's age in years
4. BMI is a number indicating the person's BMI
5. activity_score is a variable we created, and is a number that represents a person's activity level.
6. SEX is a number that indicates whether an individual is male(0) or female(1)
7. RACEA is an integer that gives a code that corresponds to an individual's race category
8. income_high is a number that indicates if an individual is high income based on our grouping (1) or not high-income (0).
9. income_middle is a number that indicates if an individual is middle income based on our grouping (1) or not middle-income (0).
10. health_binary is a character that indicates whether and individual's health is "good" or "poor"
11. WORFREQ is an integer that corresponds to a category (1-5) for how often an individual worries, 1 being daily and 5 being never.

# Visualization

```{r}
# Create scatter plot of weight versus hours of sleep
ggplot(data, aes(x = HRSLEEP, y = WEIGHT))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE)
```

Although the strength and magnitude of the fitted line are weak, the slope is negative, indicating an inverse relationship between hours of sleep and weight in pounds as we hypothesized.

```{r}
worfreq_labels <- c(
  "1" = 'Daily',
  "2" = 'Weekly',
  "3" = 'Monthly',
  "4" = 'A few times a year',
  "5" = 'Never'
)

# Facet scatterplot by worry frequency
ggplot(data, aes(x = HRSLEEP, y = WEIGHT))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE) +
  facet_grid(cols=vars(WORFREQ), labeller=labeller(WORFREQ = worfreq_labels))
```

For all worry frequencies except WORFREQ = 1 (daily), the relationship between hours of sleep and weight remains negative. For observational units who worried daily, however, the relationship between HRSLEEP and WEIGHT appears slightly positive. This hints at a possibly significant interaction between HRSLEEP/WEIGHT and a binary WORFREQ variable with 1 coded as worrying daily and 0 coded as worrying less than daily.

```{r}
race_labels <- c("100" = 'White',
              "200" = 'Black',
              "310" = 'Native A.',
              "411" = 'Chinese',
              "412" = 'Filipino',
              "416" = 'Asian Indian',
              "434" = 'Other Asian',
              "580" = 'NR')

# Facet scatterplot by race
ggplot(data, aes(x = HRSLEEP, y = WEIGHT))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE) +
  facet_grid(col=vars(RACEA), labeller=labeller(RACEA = race_labels))
```

# Final Data Wrangling

```{r}
data <- data %>% 
  mutate(worfreq_binary = ifelse(WORFREQ == 1, 1, 0)) %>% 
  mutate(race_asian = ifelse(RACEA == "411" | RACEA == "412" | RACEA == "416" |
                          RACEA == "434", 1, 0),
         race_black = ifelse(RACEA == "200", 1, 0),
         race_na = ifelse(RACEA == "310", 1, 0)) %>% 
  filter(RACEA != "580")
```

```{r}
# Facet scatterplot by worry frequency binary
ggplot(data, aes(x = HRSLEEP, y = WEIGHT))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  facet_wrap(~worfreq_binary) +
  geom_smooth(method = lm, se = FALSE)
```

Based on the previous scatterplots faceted by race, we decided to combine Asian ethnicities and keep Alaskan Native or American Indians separate. NR stands for primary race not releasable. We also recoded each categorical variable as a binary variable.

After recording worry frequency as a binary variable, we re-evaluated the relationship between sleep and weight faceted by daily vs non-daily worry frequency. We see a fairly clear difference in the two slopes of the two lines, suggesting the need for an interaction term between worry frequency and sleep in our multiple regression model.

# Checking Collinearity between Sleep and Worry Frequency

```{r}
ggplot(data, aes(x = worfreq_binary, y = HRSLEEP)) +
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE)
```

Based on the scatterplots of binary worry frequency vs hours of sleep, there does not appear to be a significant relationship between the two, and so we do not suspect collinearity between our two explanatory variables.

```{r}
# Exporting wranled dataset for data analysis use
write.csv(data, "filename.csv")
```

# Scatterplots with BMI as response

```{r, warning = FALSE, message = FALSE}
# Create scatter plot of BMI versus hours of sleep
ggplot(data, aes(x = HRSLEEP, y = BMI))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE)
```

```{r}
# Facet scatterplot by worry frequency
ggplot(data, aes(x = HRSLEEP, y = WEIGHT))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  facet_wrap(~WORFREQ) +
  geom_smooth(method = lm, se = FALSE)
```

```{r}
#Age versus BMI
ggplot(data, aes(x = AGE, y = BMI))+
  geom_point(alpha = 0.1) + 
  geom_jitter() +
  geom_smooth(method = lm, se = FALSE)
```

# Tallies after data wrangling

```{r}
# Look at distributions of variables
favstats(~BMI, data = data)
```

```{r}
favstats(~HRSLEEP, data = data)
```

```{r}
favstats(~AGE, data = data)
```

```{r}
favstats(~activity_score, data = data)
```

```{r}
tally(~SEX, data = data)
```

```{r}
tally(~income_high, data = data)
tally(~income_middle, data = data)
```

```{r}
tally(~race_asian, data = data)
tally(~race_black, data = data)
tally(~race_na, data = data)
```

```{r}
tally(~health_binary, data = data)
```

```{r}
tally(~worfreq_binary, data = data)
```
