---
title: "Final Data Appendix"
author: "Rose Porta and Clara Li"
date: "6/6/2020"
output: pdf_document
---

## Checking Regression Assumptions: 

# Linearity

The empirical logit plot seems approximately linear.

```{r, warning = FALSE, message = FALSE}
# Code from 18-simplelogistic-lab activity to generate x and y values for plot to check for linearity
data_bin <- data %>%
  # Break HRSLEEP into 10 bins
  mutate(HRSLEEP_group = cut(HRSLEEP, breaks=10)) %>% 
  group_by(HRSLEEP_group) %>%
  # calculate mean HRSLEEP value and proportion obese in each bin
  mutate(binned.y = mean(chronic), binned.x = mean(HRSLEEP)) %>% 
  # transform binned.y from proportion to log(odds)
  mutate(binned.y_logit = logit(binned.y))

# Create plot of log(odds) verus age
ggplot(data_bin, aes(x = binned.x, y = binned.y_logit)) +
  geom_point() + 
  geom_smooth(method = lm, se = FALSE)
```
## Randomness

Obviously, individuals are not assigned BMIs by a spinner model, but since we have a random sample, the randomness assumption is met.

## Independence

As established above for the linear regression model, the independence assumption is reasonable for this data set.

```{r}
data.cor = cor(data)
```
